### 2015-06-19 doc/thesis/system.tex
Back porting projections from paper to thesis

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*} =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Nil{*}$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{closeall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{closeset}.
Then, the function \emph{frv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} has type
\[
\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
\]
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division.
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Void$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S_{2}$ that should be bound
to a variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST4} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]

@@

In the rule that type checks the declaration of unannotated variables,
our type system uses the pair $(x,S_{2})$ to bound a union of
second-level types $S_{2}$ to a variable $x$ in the projection
environment $\penv$.
In our example, declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$,
where the projection variable $x$ bounds to 
\[
(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
in the projection environment $\penv$.
Rule \textsc{T-LOCAL2} illustrates this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
We can define \emph{infer} as follows:
\begin{align*}
infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
general(T_{i}) & \text{if $i < n$}\\
general(T_{n} \cup \Nil) & \text{if $i >= n$}
\end{array} \right.
\\ 
general(\False) & = \Boolean\\
general(\True) & = \Boolean\\
general({\it int}) & = \Integer\\
general({\it float}) & = \Number\\
general({\it string}) & = \String\\
general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
general(\mu x.T) & = \mu x.general(T)\\
general(T) & = T\\
\end{align*}

\begin{align*}
general2(\Void) & = \Void\\
general2(T{*}) & = general(T){*}\\
general2(T \times P) & = general(T) \times general2(P)\\
general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
\end{align*}

@@

Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
\[
fpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
gpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

>>>>

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*} =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Nil{*}$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; \lfloor f(a[``x"]) \rfloor_{0}$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{closeall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{closeset}.
Then, the function \emph{frv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} has type
\[
\Integer \times \Integer \times \Nil{*} \rightarrow (\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division.
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Nil{*}$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $E$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S)$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S$ that should be bound
to a fresh variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST4} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\;
       n = |\;\overline{e}\;| \\
       S = T_{n+1} \times ... \times T_{n+m} \times \Nil{*} \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Nil{*} \;\;\;
       x \notin \penv \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2})
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]

@@

In the rule that type checks the declaration of unannotated variables,
our type system uses the pair $(x,S)$ to bound a union of
second-level types $S$ to a variable $x$ in the projection
environment $\penv$.
In our example, declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$,
where the projection variable $x$ bounds to 
\[
(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
in the projection environment $\penv$.
Rule \textsc{T-LOCAL2} illustrates this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
       \env_{3} = \env_{2}[id_{1} \mapsto infer(E,1), ..., id_{n} \mapsto infer(E,n)] \\
       \env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4} - \{\overline{id}\})[\overline{id \mapsto \env_{2}(id)}]}
\end{array}
\]

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.

@@

Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x}\\
       S_{t} = fonilpt(\penv(x), i) \;\;\;
       S_{e} = finilpt(\penv(x), i) \\
       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2} \;\;\;
       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{3}\\
       \env_{4} = join(\env_{2}, \env_{3})
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fonilpt} and \emph{finilpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
The former function filters out the tuples that contain $\Nil$ in the $i$-th component,
while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
\[
fonilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
finilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

Typed Lua does not allow assignments to left-hand values that are bound
to a projection type.
This kind of assignment would be unsound, because it could break the
dependency relation that the components of each tuple of the union have.
For instance, the following example does not type check:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$r = ``foo";$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

In this example, the projected type of $r$ outsided of the $\mathbf{if}$
statement is $\Integer \cup \String$, so the assignment looks fine.
However, the projected type of $r$ inside the $\mathbf{if}$ branch is $\Integer$,
not matching the $\String$ value that $r$ has after the assignment.
